---
title: "SPCA tuning using IS"
author: "Elliot Kennedy"
date: "2023-12-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)  # For generating multivariate normal data
library(ggplot2)
library(elasticnet)  # Load the elasticnet package for sparse PCA
library(matrixcalc)  # Load necessary library for nearestPD function
library(Matrix)
library(dplyr)
```

Generating toy data for sparse PCA analysis:

```{r}
# Function to generate a positive definite covariance matrix
generate_positive_definite_cov_matrix <- function(n_features, sparsity) {
  # Generate a random sparse covariance matrix
  cov_matrix <- matrix(0, n_features, n_features)
  non_zero_entries <- sample(1:(n_features^2), sparsity)
  cov_matrix[non_zero_entries] <- runif(sparsity, 0.1, 0.5)
  cov_matrix <- Matrix::nearPD(cov_matrix)$mat  # Ensure positive definiteness
  return(cov_matrix)
}

# Set random seed for reproducibility
set.seed(123)

# Number of samples and features
n_samples <- 10
n_features <- 300
sparsity <- 45000  # Number of non-zero entries in the covariance matrix

# Generate a positive definite covariance matrix with sparsity
true_cov_matrix <- generate_positive_definite_cov_matrix(n_features, sparsity)

# Generate correlated data using the covariance matrix
data <- mvrnorm(n_samples, mu = rep(0, n_features), Sigma = true_cov_matrix)

data <- data.frame(data)
# Plot the correlation matrix of the generated data
cor_matrix <- cor(data)
# Convert the correlation matrix to a long format
cor_matrix_long <- reshape2::melt(cor_matrix)

# Create the heatmap plot using ggplot2
heatmap_plot <- ggplot(cor_matrix_long, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "#02d7f2", high = "#E93CAC", name = "Correlation") +
  labs(title = "Correlation Matrix of Generated Data")

# Print the heatmap plot
print(heatmap_plot)
```

The parameter we are tuning, alpha, controls an L1 penalty term. We will tune the sparse PCA using the Index of Sparseness or IS. The index of sparseness is defined as

$$IS = \mathrm{PEV}_{pca}\times \mathrm{PEV}_{sparse}\times \mathrm{PS}$$

Where $\mathrm{PEV}_{pca}, \mathrm{PEV}_{sparse}$, and $\mathrm{PS}$ are the Percentage of Explained Variance (PEV) from the selected principal components of the standard PCA, the PEV from the selected principal components of the sparse PCA, and the proportion of sparsity (number of entries equal to zero in the component loadings) respectively.

Tuning penalty term $\alpha$ using index of sparseness:

```{r}
# 1. Function to create a scree plot from data
# Use the scree plot to determin the fixed number of principal components to tune using the second function
create_scree_plot <- function(data) {
  require(ggplot2)
  
  pca_result <- prcomp(data)
  eigenvalues <- (pca_result$sdev)^2
  variance_explained <- eigenvalues / sum(eigenvalues) * 100
  scree_data <- data.frame(PC = 1:length(variance_explained), VarianceExplained = variance_explained)

  scree_plot <- ggplot(scree_data, aes(x = PC, y = VarianceExplained)) +
    geom_point(color = "steelblue", size = 3) +
    geom_line(color = "red") +
    labs(title = "Scree Plot for PCA",
         x = "Principal Component",
         y = "Percentage of Variance Explained") +
    theme_minimal()
  return(scree_plot)
}

# 2. Function to perform SPCA tuning using the index of sparseness
# n_components is the fixed number of principal components to tune on - determined by the skree plot
tune_spca <- function(data, alpha_grid, n_components) {
  require(elasticnet)

  PEV_pca <- sum(prcomp(data)$sdev[1:n_components]^2) / sum(prcomp(data)$sdev^2)
  results <- lapply(alpha_grid, function(alpha) {
    spca_model <- elasticnet::spca(data, K = n_components, para = rep(alpha, n_components), type = "predictor", sparse = "penalty")
    PEV_sparse <- sum(spca_model$pev)
    proportion_of_sparsity <- sum(spca_model$loadings == 0) / length(spca_model$loadings)
    IS <- PEV_pca * PEV_sparse * proportion_of_sparsity
    return(list(model = spca_model, IS = IS))
  })
  return(results)
}

# 3. Function to find optimal IS, alpha, and corresponding SPCA model
# Utility function for retrieving optimal results
find_optimal_spca <- function(alpha_grid, spca_results) {
  IS_values <- sapply(spca_results, function(result) result$IS)
  optimal_index <- which.max(IS_values)
  optimal_alpha <- alpha_grid[optimal_index]
  optimal_IS <- IS_values[optimal_index]
  optimal_spca_model <- spca_results[[optimal_index]]$model

  return(list(optimal_alpha = optimal_alpha, 
              optimal_IS = optimal_IS, 
              optimal_spca_model = optimal_spca_model))
}

# 4. Function to graph the results of the spca tuning
# displays a graph of IS values vs alpha grid values and labels the maximum IS
plot_spca_results <- function(alpha_grid, spca_results) {
  require(ggplot2)
  IS_values <- sapply(spca_results, function(result) result$IS)
  optimal_results <- find_optimal_spca(alpha_grid, spca_results)
  optimal_alpha <- optimal_results$optimal_alpha
  optimal_IS <- optimal_results$optimal_IS

  plot_data <- data.frame(Alpha = alpha_grid, IS = IS_values)
  spca_plot <- ggplot(plot_data, aes(x = Alpha, y = IS)) +
    geom_line(color = "blue") +
    geom_point(aes(x = optimal_alpha, y = optimal_IS), color = "red", size = 4) +
    labs(title = "SPCA Tuning: IS vs Alpha",
         x = "Alpha (Sparsity Parameter)",
         y = "Index of Sparseness (IS)") +
    theme_minimal()
  return(spca_plot)
}

scree_plot <- create_scree_plot(data)
print(scree_plot)

n_components <- 3 # based on scree plot
alpha_grid <- seq(0.05, 1, by = 0.05)
spca_results <- tune_spca(data, alpha_grid, n_components)
optimal_results <- find_optimal_spca(alpha_grid, spca_results)
spca_plot <- plot_spca_results(alpha_grid, spca_results)
optimal_results$optimal_alpha
optimal_results$optimal_IS
print(spca_plot)
```

